---
title: "Hierarchical Modeling"
format: html
---

## Simulating Homogeneous Trials

In this section, we simulate data from a set of homogeneous clinical trials.

We generate 5 trials that are similar in design and patient population:

- Each trial uses 1:1 randomization between treatment and control arms
- All trials draw patients from the same underlying population, with the same baseline severity distribution
- The treatment effect is identical across all trials

```{r}
#| message: false
#| 
library(dplyr)
library(ggplot2)
source("R/simulate_trials.R")

# Simulate 5 trials with 100 patients each (50 per arm)
df <- simulate_trials(
  n_trials = 5,
  n_per_arm = 50,
  beta_W = 0,     # no within-trial effect modification
  beta_A = 0      # no across-trial effect modification
)
```


## Naive pooling

@fig-naive-pooling visualizes the relationship between baseline severity and change from baseline within each treatment arm using a naively pooled analysis.


```{r}
#| message: false
#| echo: false
#| fig-cap: "Naive Pooled Analysis of Change from Baseline by Treatment Group"
#| label: fig-naive-pooling
#| fig-width: 10
#| fig-height: 8
library(ggplot2)
library(dplyr)

# Precompute mean labels
mean_labels <- df %>%
  group_by(trta) %>%
  summarise(mean_change = mean(change, na.rm = TRUE)) %>%
  mutate(label = paste0("Mean change: ", round(mean_change, 1)))

ggplot(df, aes(x = baseline, y = change)) +
  geom_point(aes(color=trial),alpha = 0.6) +
  geom_smooth(method = "lm", aes(group = 1), color = "black") +
  geom_text(
    data = mean_labels,
    aes(x = -Inf, y = Inf, label = label),
    hjust = -0.1, vjust = 1.2,
    inherit.aes = FALSE
  ) +
  labs(
    title = "Change from Baseline by Treatment Group",
    subtitle = "No within- or across-trial effect modification",
    x = "Baseline Severity",
    y = "Change from Baseline"
  ) +
  facet_wrap(~ trta, ncol = 1) +
  theme_minimal() +
  facet_wrap(~ trta, ncol = 1) +
  theme_minimal() +
  theme(legend.position = "bottom")

```

The plot shows that:

- The mean change in the active group is `r mean_labels %>% filter(trta == "Active") %>% pull(mean_change) %>% round(digits = 1)`
- The mean change in the control group is `r mean_labels %>% filter(trta == "Control") %>% pull(mean_change) %>% round(digits = 1)`

**Question for Discussion**: Can we interpret the pooled treatment effect as simply the difference in means?
